% GRASP: Copyright 1997,1998,1999,2000  Bruce Allen
\def\versionnumber{{\bf GRASP RELEASE 1.9.8}}
% $Id manual.tex,v 1.76 1999/06/25 00:14:01 ballen Exp $
% GRASP USERS MANUAL COPYRIGHT BRUCE ALLEN, 1997, 1998, 1999

% UNCOMMENT TO INCLUDE ONLY ONE OR MORE SECTIONS
% \includeonly{comma-separated file list (with no .tex extensions)}
% \includeonly{man_intro}
% \includeonly{man_correlation}
% \includeonly{man_40meter}
% \includeonly{man_utility}
% \includeonly{man_inspiral}
% \includeonly{man_ringdown}
% \includeonly{man_template}
% \includeonly{man_transient}
% \includeonly{man_binary-search}
% \includeonly{man_stochastic}
% \includeonly{man_frame}
% \includeonly{man_GRtoolbox}
% \includeonly{man_timefrequency}

\makeindex

\documentclass[11pt]{article}
\usepackage{times,epsfig,lgrind}
\usepackage[hypertex,pagebackref=true,letterpaper=true,linkcolor=blue]{hyperref}
\usepackage{fancyhdr}

%define page size
\setlength{\textheight}{9.0in}
\setlength{\textwidth}{6.5in}
\setlength{\topmargin}{-0.25in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}

%set up headers and footers on each page
\pagestyle{fancy}
\fancyhf{}
\lhead{\bf\begin{tabular}{c}Section\\\thesubsection\end{tabular}}
\chead{\bf\leftmark}
\rhead{\bf\begin{tabular}{c}Page\\\thepage\end{tabular}}
\lfoot{\bf\versionnumber}
\cfoot{\bf Page~\thepage}
\rfoot{\bf\today}

% Define marks without the section numbers
\renewcommand{\sectionmark}[1]{\markboth{#1}{}}
\renewcommand{\subsectionmark}[1]{\markright{#1}}

% Make equation numbers reset to zero at start of each subsection
\def\theequation{\thesubsection.\arabic{equation}}
\renewcommand{\refname}{}
\date{\today}
\makeatletter
\@addtoreset{equation}{subsection}
\makeatother
\providecommand{\eth}{\partial\!\!{}^{-}}


% Start the document
\begin{document}

\title{GRASP: a data analysis package for gravitational wave detection}
\author{Bruce Allen\thanks{\tt ballen@dirac.phys.uwm.edu}\\
Department of Physics\\
University of Wisconsin - Milwaukee\\
PO Box 413\\
Milwaukee WI 53201, USA}

\maketitle

\begin{center}
{\bf Contributors:}
\end{center}
Warren Anderson,
R. Balasubramanian,
Kent Blackburn,
Patrick Brady,
Jim Brau,
Jolien Creighton,
Teviet Creighton,
Steve Drasco,
Serge Droz,
Eanna Flanagan,
Wensheng Hua,
Scott Hughes,
Dustin Laurence,
Adrian Ottewill,
Ben Owen,
Jorge Pullin,
Joseph Romano, and
Alan Wiseman.
\vskip 1.0 in

\begin{abstract}
\noindent
GRASP ({\bf G}ravitational {\bf R}adiation {\bf A}nalysis \& {\bf
S}imulation {\bf P}ackage) is a public-domain software tool-kit designed
for analysis and simulation of data from gravitational wave detectors.
This users manual describes the use and features of this package.
Note: an up-to-date version of this manual may be obtained at:
\htmladdnormallink{{\tt http://www.lsc-group.phys.uwm.edu/$\sim$ballen/grasp-distribution/}}
{http://www.lsc-group.phys.uwm.edu/~ballen/grasp-distribution/}.
The software package is also available from this site.
\end{abstract}
\vskip -6.5 in \centerline{\bf USERS MANUAL}
\vskip 7.0 in

\begin{center}
Copyright 1999 \copyright Bruce Allen
\end{center}
\vskip 0.5in
\begin{center}
{\bf \versionnumber}
\end{center}
\clearpage
\chead{\bf TABLE OF CONTENTS }
\lhead{\bf\begin{tabular}{c}Section\\0\end{tabular}}
\tableofcontents
\clearpage
\chead{\bf \leftmark}
\lhead{\bf\begin{tabular}{c}Section\\\thesubsection\end{tabular}}

\section{ACKNOWLEDGEMENTS}
This work has been partially supported by National Science Foundation
grants to: the University of Wisconsin - Milwaukee PHY9507740 and PHY9728704, the
LIGO project PHY9210038 and the LIGO visitors program PHY9603177.
It has also benefited from the contributions of several individuals.
Alan Wiseman wrote the chirp generation routines with me in 1995, then
entirely re-wrote them for GRASP.  Joseph Romano took my (AVS versions of
the) stochastic background code, helped to fix a number of the problems,
and produced the stochastic background simulation routines contained in
GRASP.   Sathyaprakash generously provided me with a copy of his routine
{\tt grid4.f} which eventually evolved into {\tt template\_grid}.  Jolien
Creighton wrote and documented the section on black hole ringdown, and
incorporated the FFTW optimized Fourier transform routines.  
Teviet Creighton, Ben Owen, Scott Hughes and Patrick Brady contributed
essential pieces towards the binary-inspiral search performed on the
Caltech 40-meter data from November 1994. Jim Mason
was ``instrumental" in helping me to understand and document the data
format used in the 40-meter prototype experiment.  Fred Raab, Bob Spero,
Stan Whitcomb, Kent Blackburn and others have contributed many useful
ideas and insights about how to understand the data stream from a real
interferometer.  Kip Thorne and his research group provided or called
attention to many of these ideas.
\clearpage

\chead{\bf\begin{tabular}{c}\leftmark\\\rightmark\end{tabular}}
\include{man_intro}
\include{man_40meter}
\include{man_frame}
\include{man_correlation}
\include{man_inspiral}
\include{man_testmass}
\include{man_ringdown}
\include{man_template}
\include{man_timefrequency}
\include{man_stochastic}
\include{man_galaxy}
\include{man_binary-search}
\include{man_transient}
\include{man_periodic}
\include{man_utility}
\include{man_GRtoolbox}
%%%   References
%
\clearpage
\chead{\bf REFERENCES}
\section{REFERENCES}
\begin{thebibliography}{99}

\bibitem{NumRec}
W.H. Press, B.P. Flannery, S.A. Teukolsky and W.T. Vettering, {\it
Numerical recipes in C: the art of scientific computing}, 2nd edition,
Cambridge University Press.

\bibitem{MPI}
Message Passing Interface Forum, {\it MPI: a message passing interface
standard}, International Journal of Supercomputer Applications {\bf 8} 3/4 1994.

\bibitem{MPICH}
W. Gropp and E. Lusk, {\it User's Guide for mpich, a portable implementation
of MPI}, Technical Report ANL/MCS-TM-ANL-96/6, Argonne National Laboratory.

\bibitem{Bala}
R. Balasubramanian, B.S. Sathyaprakash, and S.V. Dhurandhar, {\it
Gravitational waves from coalescing binaries:  detection strategies and
monte carlo estimation of parameters}, Phys.  Rev.  {\bf D53} (1996)
3033-3055; Erratum in Phys. Rev. {\bf D54} (1996) 1860.  Originally
posted as gr-qc/9508011.

\bibitem{Owen}
B.J. Owen, {\it Search templates for gravitational waves from
inspiraling binaries:  choice of template spacing}, Phys. Rev. {\bf
D53} (1996) 6749-6761.  Originally posted as gr-qc/9511032.

\bibitem{OwenSathya}
B.J. Owen and B.J. Sathyaprakash, {\it Matched filtering of gravitational
waves from inspiraling compact binaries: Computational cost and template
placement}, gr-qc/9808076.

\bibitem{biww}
L. Blanchet, B.R. Iyer, C.M. Will, and A.G. Wiseman,
{\it Gravitational waveforms from inspiralling compact binaries
to second-post-Newtonian order},
Class. Quantum Grav. {\bf 13}, 575-584 (1996).

\bibitem{willwiseman}
C.M. Will and A.G. Wiseman, {\it Gravitational radiation from compact
binary systems: Gravitational waveforms and energy loss to second
post-Newtonian order}, Phys. Rev. {\bf D54} (1996) 4813-4848.

\bibitem{blanchet:1996}
L. Blanchet, Phys.\ Rev.\  {\bf D54}, 1417 (1996).

\bibitem{cutleretal}
C. Cutler {\it et al.}, {\it The Last Three Minutes: Issues in 
Gravitational-Wave Measurements of Coalescing Binaries},
Phys. Rev. Lett. {\bf 70} (1993) 2984-2987.

\bibitem{lincolnwill}
C.W. Lincoln and C.M. Will, {\it Coalescing binary systems of compact
objects to (post)$^{5/2}$-Newtonian order: Late-time evolution and
gravitational-radiation emission}, Phys. Rev. {\bf D42} (1990) 1123-1143.

\bibitem{bdiww}
L. Blanchet, T. Damour, B.R. Iyer, C.M. Will, and A.G. Wiseman,
{\it Gravitational-Radiation Damping of Compact Binary Systems
to Second Post-Newtonian Order},
Phys. Rev. Lett. {\bf 74} (1995) 3515-3518.

\bibitem{flanhughes1}
E. Flanagan and S. Hughes, {\it Measuring gravitational waves from binary
black hole coalescences.  I. Signal to noise for inspiral, merger,
and ringdown}, Phys. Rev. {\bf D57}, 4535-4565, (1998).

\bibitem{flanhughes2}
E. Flanagan and S. Hughes, {\it Measuring gravitational waves from binary
black hole coalescences.  II. The waves' information and its extraction,
with and without templates}. Phys. Rev. {\bf D57}, 4566-4587, (1998).

\bibitem{finnandchernoff}
L. S. Finn and D. Chernoff,  {\it Observing binary inspiral in gravitational
radiation: one interferometer}. Phys. Rev. {\bf D47}, 2198-2219, (1993).

\bibitem{thornetofinnhughes}
K. S. Thorne, Private communication to Scott Hughes.

\bibitem{thornefinnerror}
Thorne's estimate of 1.04 is off by 6\% because it uses a fit to the
probability distribution $P(\Theta^2 > x^2)$ (defined in
{\cite{finnandchernoff}}) which is not accurate enough to reliably
compute the $x^2$ moment of that distribution.

\bibitem{markovic}
D. Markovi\'c, {\it Possibility of determining cosmological parameters
from measurements of gravitational waves emitted by coalescing compact
binaries}, Phys. Rev. {\bf D48} 4738-4756 (1993).

\bibitem{dwhogg}
D.W. Hogg, notes, {\it Manual on Cosmological Distance Measures},
available from
\htmladdnormallink{{\tt http://www.sns.ias.edu/$\sim$hogg}.}
{http://www.sns.ias.edu/~hogg}

\bibitem{kolbturner}
E.W. Kolb and M.S. Turner,  {\it The Early Universe},
published by Addison Wesley, 1990.

\bibitem{cutler:1994}
C. Cutler and {\'E}.E. Flanagan \emph{Gravitational waves from
  merging compact binaries: How accurately can one extract the binary's
  parameters from the inspiral waveform?} Phys. Rev. \textbf{D49}
  2658--2697 (1994).

\bibitem{droz:1999}
S.Droz, D.J. Knapp, E. Poisson, B.J. Owen available at\\
\htmladdnormallink{{\tt http://xxx.lanl.gov/abs/gr-qc/9901076}.}
{http://xxx.lanl.gov/abs/gr-qc/9901076}

\bibitem{poisson:1995}
E. Poisson and C.~M. Will, Phys.\ Rev.\  {\bf D52}, 848 (1995).

\bibitem{matthewsandwalker}
J. Mathews and R.L. Walker, {\it Mathematical Methods of Physics},
Second Edition, Addison-Wesley, 1970.

\bibitem{echeverria:1989}
F. Echeverria \emph{Gravitational-wave measurements of the mass
  and angular momentum of a black hole,} Phys. Rev. \textbf{D40}
  3194--3203 (1989).

\bibitem{goldberg:1967}
J.N. Goldberg, A.J. Macfarlane, E.T. Newman, F. Rohrlich, and E.C.G. Sudarshan
  \emph{Spin-$s$ spherical harmonics and $\eth$,} J. Math. Phys.
  \textbf{8} 2155--2161 (1967).

\bibitem{leaver:1985}
E.W. Leaver \emph{An analytic representation for the quasi-normal
  modes of {K}err black holes,} Proc. Roy. Soc. Lond. \textbf{A402} (1985).

\bibitem{onozawa:1997}
H. Onozawa \emph{Detailed study of quasinormal frequencies of the
  {K}err black hole,} Phys. Rev. \textbf{D55} 3593--3602 (1997).

\bibitem{press:1973}
W.H. Press and S.A. Teukolsky (Oct 1973) \emph{Perturbations of a rotating
  black hole. {II}. {D}ynamical stability of the {K}err metric,}
  Astrophys. J. \textbf{185} 649--673 (1973).

\bibitem{teukolsky:1973}
S.A. Teukolsky \emph{Perturbations of a rotating black hole. {I}.
  {F}undamental equations for gravitational, electromagnetic, and
  neutrino-field perturbations,} Astrophys. J. \textbf{185} 635--647 (1973).

\bibitem{Gillespie}
A.D. Gillespie, {\it Thermal Noise in the Initial LIGO Interferometers},
Caltech PhD thesis, 1995.

\bibitem{Lyons}
T.T. Lyons, {\it An optically recombined laser interferometer for
gravitational wave detection}, Caltech PhD thesis, 1997.

\bibitem{timefreqpap}
Warren G. Anderson and R. Balasubramanian, {\em Time-Frequency Detection
of Gravitational Waves},
Submitted for publication,
\htmladdnormallink{{\tt
http://xxx.lanl.gov/abs/gr-qc/9905023/}}
{http://xxx.lanl.gov/abs/gr-qc/9905023/}.

\bibitem{Stegerscode}C. Steger, {\em {\rm computer code} {\sc detect-lines 1.2}},
Technische Universitat M\"{u}nchen, M\"{u}nchen Germany, 1996, available from\\
\htmladdnormallink{{\scriptsize \tt ftp://ftp9.informatik.tu-muenchen.de/pub/detect-lines/detect-lines-1.2.tar.gz}}
{ftp://ftp9.informatik.tu-muenchen.de/pub/detect-lines/detect-lines-1.2.tar.gz}.

\bibitem{Stegerspap} C. Steger, IEEE Transactions on Pattern Analysis
and Machine Intelligence, {\bf 20}, 113, (1998).

\bibitem{AllenReview}
B.~Allen, {\it The stochastic gravity-wave background: sources and
detection}, in Relativistic Gravitation and Gravitational Radiation,
Proceedings of the Les Houches School on Astrophysical
Sources of Gravitational Radiation, eds. J-A. Marck and J-P. Lasota,
(Cambridge University Press, Cambridge, England, 1997) pages 373-417.

\bibitem{wortz}
See equations (4.13) and (4.14) in
{\it Spacecraft attitude determination and control}, 
Ed.~James R.~Wortz,
(D.~Reidel Publishing Co., Boston, 1985).

\bibitem{Science92}
A.~Abramovici et al., 
{\it Science} {\bf 256}, 325 (1992). 

\bibitem{thomson82}
D.J. Thomson, {\it Spectrum estimation and harmonic analysis}, 
Proceedings of the IEEE, {\bf 70}, 1055-96, (1982).

\bibitem{percivalwalden}
D.B. Percival and A.T. Walden, {\it Spectral analysis for physical applications}, first edition, Cambridge University Press, (1993).

\bibitem{parklees}
J.M. Lees and J. Park, {\it Multiple-taper spectral analysis: A stand-alone C
subroutine}, Computers and Geology {\bf 21}, 199-236.

\bibitem{LS95}
D. Lai and S. L. Shapiro, {\it Astrophys. J.} {\bf 442}, 259-272, (1995).

\bibitem{Chandra69}
S. Chandrasekhar, {\it Ellipsoidal Figures of Equilibrium}, (Yale University
Press, New Haven, 1969).

\bibitem{LaiPC}
Dong Lai, Personal Communication.

\bibitem{ericsfiles} The data files for the amplitueds $A_{lm}(v)$ and the
luminosity function $P(v)$ were kindly provided by Eric Poisson
(poisson@physics.uoguelph.ca). 

\bibitem{droz97} S. Droz and E. Poisson, {\it Gravitational waves from
inspiraling compact binaries: Second post Newtonian waveforms as search
templates}, Phys.\ Rev.\  {\bf D57}, 4449 (1997).

\bibitem{frame} Benoit Mours, {\it Frame Library User's Manual},
version 3.60, \htmladdnormallink{{\tt
http://wwwlapp.in2p3.fr/virgo/FrameL/}}
{http://wwwlapp.in2p3.fr/virgo/FrameL/}, email: {\tt mours@lapp.in2p3.fr},
VIR-MAN-LAP-5400-103.

\bibitem{40msearch}
B.~Allen, K.~Blackburn, P.~Brady, J.~Creighton, T.~Creighton, S.~Droz, A.~Gillespie,
S.~Hughes, S.~Kawamura, T.~Lyons, J.~Mason, B.J.~Owen, F.~Raab, M.~Regehr, B.~Sathyaprakash,
R.L.~Savage, Jr., S.~Whitcomb, A.G.~Wiseman,
{\it Observational limit on gravitational waves from binary neutron stars in the Galaxy},
Phys.\ Rev.\ Lett.\ {\bf 83}, 1498 (1999).  Also
available from:
\htmladdnormallink{{\tt http://xxx.lanl.gov/abs/gr-qc/9903108}}{http://xxx.lanl.gov/abs/gr-qc/9903108}.

\bibitem{ABW}
Anninos, Brandt, and Walker, {\it New orthogonal body-fitting coordinates
for colliding black-hole spacetimes},  Phys. Rev. D \textbf{57}
6158--6167 (1998).

\end{thebibliography}
\index{alastpage}
\end{document}
